---
title: "Introduction to R"
author: "Student Name, Student ID, StudentName@uni-muenster.de"
subtitle: Exam Summer Term 2018
output:
  html_notebook:
    code_folding: show
    number_sections: yes
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
    number_sections: yes
    keep_tex: true
---

***

* Answer **7** out of **10** of the following exercises in either German or English.

* Hand in your solutions before Friday, 13 April 2018 at 10 am.

* It is advised to regularly check the learnweb and your emails in case of urgent updates.

* Please sent your solutions files to [Willi Mutschler](mailto:willi.mutschler@wiwi.uni-muenster.de?subject=Exam Intro R). We will confirm the receipt of your work also by email.

* The solution files should contain your executable and commented script file or preferably a R Notebook. 

* You may use _any_ available R package you find fit to solve the exercise.

* Please label your axes and title in your plots.

* **All students must work on their own.**

***

\pagebreak

# Linear Equation
Solve the linear equation $A\cdot x = b$ with
$$
A = \begin{pmatrix} 1 & 1 & 1 & 3\\ 2 & 5 & 8 & 6\\ 4 & 3 & 7 & 9\\ 3 & 6 & 5 & 1 \end{pmatrix} \text{ and }
b = \begin{pmatrix} 1 \\ 4 \\ 3 \\ 5 \end{pmatrix}
$$

***Solution:***
```{r}
x <- "2"
print(x)
```
and compute the inverse as well as Eigenvalues of $A$.

***Solution:***
```{r}

```
\pagebreak

***
# Functions

1. Write a function `psum(n,a)` that computes
$$
s_{n,a} := \sum_{k=0}^n \frac{k^a}{k^a+1}
$$
for any natural number $n \in \{1,2,...\}$ and any $a > 0$.

***Solution:***
```{r}

```

2. Write a function `mymatrix(n)` that returns a $n \times n$ matrix such that: the first and last row as well as the first and last column contain only ones, whereas the remaining values are zero, e.g. 
$$
\begin{pmatrix} 1 & 1 & 1 & 1\\ 1&0&0&1\\1&0&0&1\\1 & 1 & 1 & 1
\end{pmatrix}
$$

***Solution:***
```{r}

```
\pagebreak

***
# Passenger numbers
The file **apass.csv** contains monthly data on passenger numbers of US airlines from January 1949 to December 1959. 

1. Read the data into a data frame.

***Solution:***
```{r}

```

2. Create a vector that contains the corresponding dates. Add this vector to your data frame.

***Solution:***
```{r}

```

3. Plot the passenger numbers against the date vector.

***Solution:***
```{r}

```

4. Calculate the mean passenger numbers for each month. Plot the means as a bar chart.

***Solution:***
```{r}

```
\pagebreak

***
# Porsche
The file **Porsche911.csv** contains data on Porsche 911 cars, `name` is an id for the owner, `loc` is the location, `age` is the age of the car, `TKM` is the mileage in thousands kilometers and `price` the current price listed on an internet plattform for used cars in thousand Euros.

1. Read in the data into a data frame and compute key descriptive statistics (mean, standard deviation, smallest and largest values, quartiles, covariance and correlation) for the variables `age`, `TKM` and `price`.

***Solution:***
```{r}

```

2. Plot boxplots for `age`, `TKM` and `price`.

***Solution:***
```{r}

```

3. Generate the empirical cumulative distribution function as well as histograms for each of the variables `age`, `TKM` and `price`.

***Solution:***
```{r}

```
\pagebreak

***
# Law of large numbers

Let $X_1, X_2,...$ be a sequence $X_1,X_2,...$ from an $AR(1)$ process:
$$
\left( X_{i}-\mu \right) =\rho \left( X_{i-1}-\mu \right) +\varepsilon _{i}
$$
where $\varepsilon _{i}$ is uniformly distributed on the interval $[-1,1]$ and $|\rho |<1$. Define the sequence of random variables
$$
\bar{X}_{n}=\frac{1}{n}\sum_{i=1}^{n}X_{i}.
$$
The weak law of large numbers states that the sample average $\bar{X}_{n}$ converges in probability towards the expected value $\mu$ when $n$ tends to infinity.

Show by simulation that the law of large numbers holds despite the intertemporal dependence in $X$. In particular, show the convergence by means of an appropriate plot. Hint: You may set the parameters to e.g. $\rho=0.5$ and $\mu=2$ (or any other value you find fit.)

***Solution:***
```{r}

```
\pagebreak


***
# Limits of maxima
Let $X_{1}, X_{2},...$ be an i.i.d. sequence of random variables
uniformly distributed on the interval $[0,1]$. Define the random variables
$$
M_{n}=\max_{i=1,\ldots ,n}X_{i}
$$
and its normalized version $\overline{M}_{n}=(M_{n}-1)\cdot n$. One can show that the limit distribution of $\overline{M}_{n}$ is the Weibull distribution with density $\exp \left( x\right)$. Write an R program to illustrate that $\overline{M}_{n}$ converges in distribution. To this end, set $n=100$ and $R=1000$ and consider the $R \times n$ matrix with uniformly distributed random variables $X_i$. Show the convergence by means of an appropriate plot.

Note: Try to avoid using loops (use e.g. apply instead). You will not get full points if you use a loop.

***Solution:***
```{r}

```
\pagebreak


***
# Student teacher ratio

Load the dataset **caschool.csv** into the object `caschool`. This dataset is discussed in great detail in the textbook of Stock and Watson. 

1. Make the following variables accessible:

* test score `testscr`

* student-teacher ratio `str`

* percentage of English language learners `el_pct`

* expenditures per student `expn_stu`

***Solution:***
```{r}

```
 
2. Regress `testscr` on a constant and `str`. Assign the residuals of the regression into the variable `r1`. Now regress `testscr` on an intercept, `str`, `el_pct` and `expn_stu`. Put the residuals into the variable `r2`. Compute the sum of squared residuals for both regressions. Display `r1` and `r2` in one plot, where the points of `r2` are marked red.

***Solution:***
```{r}

```

3. Consider the regression of `testscr` on a constant, `str`, `elp_ct` and `expn_stu`. Predict the value of `testscr` for a school district with an average class size (`str`) of 25 students, a percentage of English learners (`el_pct`) of 60% and an average expenditures per student (`expn_stu`) of 4000$.

***Solution:***
```{r}

```


4. Reconsider the regression of `testscr` on a constant, `str`, `el_pct` and `expn_stu`. Compute the heteroscedastic robust standard errors.

***Solution:***
```{r}

```

5. Test the null hypothesis that the coefficients on `str` and `expn_stu` both equal $0$ and the coefficient on `el_pct` equals $-0.7$. Hint: Use the linearHypothesis function of the `car` package.

***Solution:***
```{r}

```
\pagebreak

***
# Asymptotic normality

Consider the multiple linear regression model $y=X\beta +u$. In R, generate the matrix $X$ by executing the following commands:
```{r}
library(MASS)
X <- cbind(1,mvrnorm(n=100,c(5,10),matrix(c(1,0.9,0.9,1),2,2)))
```

Assume that the true coefficient vector is
$$
\beta =\left(
\begin{array}{c}
3 \\
2 \\
-1
\end{array}
\right)
$$
and the error terms are i.i.d. uniformly distributed on the interval $[-1,1]$. Hence, the assumption of normally distributed error terms is violated.

1. Write an R program that generates $R=10000$ random samples of size $n=100$ each. Generate an empty vector `Z <- rep(NA,R)`. For each
sample $i=1,\ldots ,R$, compute the OLS estimate $\hat{\beta}$ of $\beta$ and store the second component of $\hat{\beta}$ in the $i$-th element of the vector `Z`.

***Solution:***
```{r}

```

2. Plot the histogram of `Z`. Compute the mean $m$ and standard deviation $s$ of `Z` and add the density of $N(m,s)$ to the plot.

***Solution:***
```{r}

```
\pagebreak


***
# Stochastic frontier analysis

Consider the Cobb-Douglas production function
$$
y=Ax_{1}^{\alpha }x_{2}^{\beta }
$$
By definition, the production function returns the maximal output for given inputs, and actual production cannot be larger than $y$. Due to
inefficiencies, actual production could be modeled (in logs) as
$$
\ln y=\ln A+\alpha \ln x_{1}+\beta \ln x_{2}-u
$$
where $u$ is a **non-negative** random variable. Since other disturbances (e.g. measurement errors) can enter the production function, it is more common to add another, **symmetrically distributed**, disturbance term $v$,
$$
\ln y=\ln A+\alpha \ln x_{1}+\beta \ln x_{2}-u+v
$$
Assume that $u$ is exponentially ($u \sim Exp(\lambda)$) and $v$ normally ($v\sim N(0,\sigma^{2})$) distributed. One can show that if $u$ and $v$ are independent then the density function of $\varepsilon =v-u$ is given by
$$
f_{\varepsilon} (\varepsilon) =\lambda \exp\left(\lambda \varepsilon + \frac{1}{2} \lambda^{2} \sigma^{2}\right) \Phi \left( \frac{-\varepsilon}{\sigma} -\lambda \sigma \right)
$$
where $\Phi$ is the distribution function (`pnorm`) of $N(0,1)$ and $\exp$ the exponential function (`exp`).

1. Load the dataset **sfa.csv**. This dataset is an abbreviated version of table F7.2 of Greene, 2008. The original data appeared in Zellner and Revankar, _Generalized Production Functions_, Review of Economic Studies, 36 (1969), 241-250.

***Solution:***
```{r}

```

2. Write an R program to estimate the parameters $A$, $\alpha$, $\beta$, $\lambda$ and $\sigma$ by maximum likelihood on this dataset.

***Solution:***
```{r}

```

3. Compute the asymptotic standard errors.

***Solution:***
```{r}

```
\pagebreak


***
# Variance estimation in GARCH

When one considers an iid sample $X_{1},\ldots ,X_{n}$ from $X\sim N(\mu ,\sigma ^{2})$ then one usually estimates the variance $\sigma^{2}$ using
$$
\hat{\sigma}^{2}=\frac{1}{n-1}\sum_{i=1}^{n}(X_{i}-\bar{X})^{2}
$$
The distribution of the normalized estimator for the variance is given by:
$$
\frac{\left( n-1\right) \hat{\sigma}^{2}}{\sigma ^{2}}\sim \chi _{n-1}^{2}
$$
where $\sigma^2$ is the true variance. Consider the case when the observations are not iid but are stochastically dependent over time. To this end, assume that $X_{1},\ldots ,X_{n}$ is a time series generated by a $GARCH(1,1)$ process
\begin{align*}
X_{i} &\sim N(0,\sigma _{i}^{2}) \\
\sigma _{i}^{2} &=\omega +\alpha X_{i-1}^{2}+\beta \sigma _{i-1}^{2}
\end{align*}
with $\omega =0.1$, $\alpha =0.1$, $\beta =0.85$ and sample size equal to $n=2500$. Show by simulations that the distribution of the normalized estimator for the variance is not $\chi^2_{n-1}$-distributed. Hint: The true unconditional variance of this GARCH process is $\sigma^{2}=2$.

***Solution:***
```{r}

```